A Formal Model of Checked C

### 2022-08-07 sandp23/background.tex
update.

Development of Checked C was initiated by Microsoft Research in 2015 but starting in
late 2021 was forked and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the formalism~\cite{li22checkedc}.

@@

\myparagraph{Spatial Safety and Backward Compatibility}
%
\checkedc is backward compatible with legacy C in the sense that all
legacy code will type-check and compile. However, only code that
appears in \emph{checked code regions}, which we call \emph{checked code},
is spatially safe. Checked regions can be designated at the level of
files, functions, or individual code blocks, the first with a
\code{#pragma} and the latter two using the \code{checked}
keyword.\footnote{You can also designate \emph{unchecked} regions
  within checked ones.}  Within checked regions, both
legacy pointers and certain unsafe idioms (e.g., \emph{variadic} function
calls) are disallowed. The code in Fig.~\ref{fig:checkedc-example}
satisfies these conditions, and will type-check in a checked region.

How should we think about code that contains both checked and legacy
components? \citet{li22checkedc} proved, for a simple
formalization of \checkedc, that \emph{checked code cannot be blamed}:
Any spatial safety violation is caused by the execution of unchecked
code.

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance, as shown below, Checked C programs cannot use address-taken variables in a bounds expression as the bounds relations may not hold because of possible modifications through pointers.
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
Consequently, converting existing C programs to Checked C might require refactoring, e.g., modifying the above program to not use~\inlinecode{&n} expression, which might require considerable effort~\cite{duanrefactoring} depending on the program's complexity. 
Recently, Machiry et al. developed~\threec~\cite{machiry2022c} that tries to automatically convert a program to Checked C by adding appropriate pointer annotations.
However, as described in \threec, complete automated conversion is infeasible and requires the developer to convert some code regions manually.
Although, the backward compatibility of Checked C helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).

@@

\myparagraph{Maintaining Non-crashing}
Previously, the main guarantee of \checkedc \cite{li22checkedc} was the blame theorem.
The sources of crashing in \checkedc are (1) unchecked regions crash themselves;
  and (2) the misuse of checked pointers in unchecked regions.
For example, at \Cref{fig:checkedc-example-3} line 31,
we call a unchecked function \code{f} with a checked null-terminated array (NT-array) pointer argument.
At line 8, depending on the NT-array size, \code{free(s[10])} might crash.
Even if it does not crash, line 38 is doomed because of the \code{free} call.

Enlightened by program partitioning mechanism,
we sandbox the unchecked code regions and utilize the \checkedc type system
to disallow checked pointers to be used in an unchecked code region. 
To achieve the communication between checked and unchecked code regions,
Tainted pointers can be shared by different regions,
whose data are stored in the sandboxed unchecked heap.
Users are required to copy checked data to tainted pointers before they are shared in unchecked regions.
For example,  we copy the checked pointer data to the tainted pointer \code{tp}
at \Cref{fig:checkedc-example-3} line 26,
and input the tainted pointer to the unchecked function at line 34.
At line 8, even if statement might crash, since tainted pointers are stored
in the sandboxed heap, it can be recovered.
At line 37, the use of a tainted pointer in a checked region
requires a verification on it.
This is handled by inserting additional checks
and creating exception handling before the use by the \systemname compiler.
Thus, the checked pointer \code{p} is safely used at line 38.

@@

\Cref{fig:checkedc-example-1} defines a string sorting algorithm
depending on the input function pointer \code{cmp} that provides the generic order for strings,
and \code{compare_1} is an example \code{cmp} function
that adds the ASCII numbers of characters in the two strings and compare the results.
In addition, function pointers enable the callback mechanism,
i.e., a server sends a function pointer to a client in an unchecked region,
and allows the client to access some server resources by calling back the pointer.
This is a common usage between a web-browser and untrusted third party libraries.
The function call in \Cref{fig:checkedc-example-3} line 34 is one such usage.

@@

To prevent the checked pointer address leak,
we prevent any unchecked regions acknowledge checked pointer variables.
In addition, to facilitate checked function callbacks, 
the \systemname compiler compiles every checked function
with an additional tainted shell function.
Users are required to serve unchecked regions with the tainted shell pointer instead of the original checked function pointer.
For example, \code{tainted_compare_1} and \code{tainted_stringsort} in \Cref{fig:checkedc-example-2} 
are the tainted shells of the checked functions \code{compare_1} and \code{stringsort}, respectively.
In the tainted shells, the arguments are tainted versions of the corresponding arguments in the checked functions.
Inside the shell body, we create checked pointer copies of the tainted arguments, and call the checked functions.
In addition, once the checked function returns,
if the output is a checked pointer, we copy its the data to a new tainted pointer and return it.
\Cref{fig:checkedc-example-3} line 42 is an example of serving the function call living in an unchecked region
with a tainted shell pointer argument \code{tainted_compare_1}.
Even if $g$ derndomizes its address (line 14), the shell address is in the sandbox and has no harm,
and calling the shell never exposes any checked pointer information outside of the shell.

>>>>

\checkedc Development began in 2015 by Microsoft Research, but it was forked
in late 2021 and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the formalism~\cite{li22checkedc}.

@@

\myparagraph{Spatial Safety and Backward Compatibility}
%
\checkedc is backward compatible with legacy C in the sense that all
legacy code will type-check and compile. However, only code that
appears in \emph{checked code regions}, which we call \emph{checked code},
is spatially safe. Checked regions can be designated at the level of
files, functions, or individual code blocks, the first with a
\code{#pragma} and the latter two using the \code{checked}
keyword.\footnote{You can also designate \emph{unchecked} regions
  within checked ones.}  Within checked regions, both
legacy pointers and certain unsafe idioms (e.g., \emph{variadic} function
calls) are disallowed. The code in Fig.~\ref{fig:checkedc-example}
satisfies these conditions and will type-check in a checked region.

How should we approach code that has both checked and legacy
components? \citet{li22checkedc} proved, for a simple
formalization of \checkedc, that \emph{checked code cannot be blamed}:
Any spatial safety violation is caused by the execution of unchecked
code.

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance, as shown below, Checked C programs cannot use address-taken variables in a bounds expression as the bounds relations may not hold because of possible modifications through pointers.
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
Consequently, converting existing C programs to Checked C might require refactoring, e.g., modifying the above program to not use~\inlinecode{&n} expression, which might require considerable effort~\cite{duanrefactoring} depending on the program's complexity. 
Recently, Machiry et al. developed~\threec~\cite{machiry2022c} that tries to automatically convert a program to Checked C by adding appropriate pointer annotations.
However, as described in \threec, complete automated conversion is infeasible and requires the developer to convert some code regions manually.
Although, the backward compatibility of Checked C helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (i.e., checked, or safe regions).

@@

\myparagraph{Maintaining Non-crashing}
Previously, the main guarantee of \checkedc \cite{li22checkedc} was the blame theorem.
The sources of crashing in \checkedc are (1) unchecked regions crash themselves;
  and (2) the misuse of checked pointers in unchecked regions.
For example, at \Cref{fig:checkedc-example-3} line 31,
we called an unchecked function \code{f} with a checked null-terminated array (NT-array) pointer argument.
At line 8, depending on the NT-array size, \code{free(s[10])} might crash.
Even if it does not crash, line 38 is doomed because of the \code{free} call.

Enlightened by program partitioning mechanism,
we sandbox the unchecked code regions and utilize the \checkedc type system
to disallow checked pointers to be used in an unchecked code region. 
To achieve the communication between checked and unchecked code regions,
Tainted pointers can be shared by different regions,
whose data are stored in the sandboxed unchecked heap.
Users are required to copy checked data to tainted pointers before they are shared in unchecked regions.
For example,  we copy the checked pointer data to the tainted pointer \code{tp}
at \Cref{fig:checkedc-example-3} line 26,
and input the tainted pointer to the unchecked function at line 34.
At line 8, even if statement might crash, since tainted pointers are stored
in the sandboxed heap, it can be recovered.
At line 37, the use of a tainted pointer in a checked region
requires a verification on it.
This is managed by inserting additional checks
and creating exception handling before the use by the \systemname compiler.
Thus, the checked pointer \code{p} is safely used at line 38.

@@

\Cref{fig:checkedc-example-1} defines a string sorting algorithm
depending on the input function pointer \code{cmp} that provides the generic order for strings,
and \code{compare_1} is an example \code{cmp} function
that adds the ASCII numbers of characters in the two strings and compare the results.
In addition, function pointers enable the callback mechanism,
i.e., a server sends a function pointer to a client in an unchecked region,
and allows the client to access some server resources by calling back the pointer.
This is a common usage between a web-browser and untrusted third-party libraries.
The function call in \Cref{fig:checkedc-example-3} line 34 is one such usage.

@@

we prevent any unchecked regions from acknowledging checked pointer variables to avoid the checked pointer address leak.
In addition, to facilitate checked function callbacks, 
the \systemname compiler compiles every checked function
with an additional tainted shell function.
Users are required to serve unchecked regions with the tainted shell pointer instead of the original checked function pointer.
For example, \code{tainted_compare_1} and \code{tainted_stringsort} in \Cref{fig:checkedc-example-2} 
are the tainted shells of the checked functions \code{compare_1} and \code{stringsort}, respectively.
In the tainted shells, the arguments are tainted versions of the corresponding arguments in the checked functions.
Inside the shell body, we create checked pointer copies of the tainted arguments, and call the checked functions.
In addition, once the checked function returns,
if the output is a checked pointer, we copy its data to a new tainted pointer and return it.
\Cref{fig:checkedc-example-3} line 42 is an example of serving the function call living in an unchecked region
with a tainted shell pointer argument \code{tainted_compare_1}.
Even if $g$ derandomizes its address (line 14), the shell address is in the sandbox and has no harm,
and calling the shell never exposes any checked pointer information outside of the shell.

### 2022-08-07 sandp23/introduction.tex
update DynCast.
update.

\section{Introduction}

Vulnerabilities due to memory corruption, especially spatial memory corruptions, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite a large body of work that tries to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts---including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012}---have
explored means to compile C programs 
to automatically enforce spatial safety. These
approaches all impose performance overheads deemed too high for
deployment use. 
Recently, \citet{Elliott2018,li22checkedc} introduced and formalized \checkedc, an
open-source extension to C with new types and
annotations whose use can ensure a program’s spatial safety.
Importantly, \checkedc supports development that is 
incremental and compositional. Code regions (e.g.,
functions or whole files) designated as \emph{checked} enforce
spatial safety in a manner preserved by composition with
other checked regions. But not all regions must be checked: Checked
C's annotated \emph{checked pointers} are binary-compatible with legacy pointers, and
may coexist in the same code, which permits a deliberate (and
semi-automated) refactoring process. Parts of the FreeBSD kernel have
been successfully ported to \checkedc~\cite{duanrefactoring}, and overall, performance
overhead seems low enough for practical deployment.

@@

The guarantee \checkedc provides focuses on checked code regions,
where any stuck (i.e., ill-defined)
state reached by a well-typed program amounts to a spatial safety
violation; such a state can always be attributed to, i.e., 
\emph{blamed on}, the execution of code that is not in a checked core
region. But what are the guarantees of unchecked (unsafe) code regions?
Programmers might not expect to convert every code to \checkedc,
but leave some functions in unsafe regions,
in which case the unsafe region executions should not affect other parts.
In this paper, we cover this gap by introducing \systemname,
combining \checkedc with program partitioning mechanism, 
making three main contributions.

\myparagraph{Executions Cannot Crash}
Our first contribution is a clear program partition between checked and unchecked code regions,
and ensures that a \systemname program execution \textit{cannot crash} due to spatial safety violations.
We restrict the checked and unchecked pointers usage only in checked and unchecked code regions, respectively;
and develop tainted pointers for the communication between the two regions.
Heap is partitioned into two parts: a \textit{checked memory region} containing all checked pointers
 and an \textit{unchecked memory region} that is sandboxed.
Tainted pointers live in the unchecked memory region and is verified every time when using.

\myparagraph{Formaling the Type System, Semantics and Compiler}
%
Our second contribution is a core formalism named \lang, which extends
\citet{li22checkedc} with the non-crashing guarantee and other new features below.
We prove formally the \textit{non-crashing theorem}, i.e., 
a well-typed \lang program can never crash due to spatial safety violations.
We maintain the compiler formalism with the new features extended in \systemname.
Especially, we maintain the bound-check insertions for array accesses 
without ``fat'' pointer compilation scheme, while keeping the non-crashing guarantee.
To certify the simulation relation between the \lang semantics and the compiler,
we utilize the model-based randomized testing.
This is done by a conversion tool that converts expressions from \lang into
actual \checkedc code that can be compiled by the Clang \checkedc
compiler. We build a random generator of programs largely based on the
typing rules of \lang and make sure that, both statically and
dynamically, \lang and Clang Checked C are consistent after
conversion.  
To the best of our knowledge, 
\lang is the first language and compiler formalism with the program partitioning mechanism.

\myparagraph{Supporting Checked Pointer Callbacks and No Checked Pointer Exposure}
Our third contribution is an added up feature
to support checked (function) pointer callbacks in unchecked code regions.
In implementing a multi-threaded system, users might want 
to provide third party interface permitting third parties 
to write implement some new program features, while keeping those programs in unchecked code regions.
Moreover, they do want to provide them a (function) pointer pointing to checked data fields.
To the best of our knowledge, \systemname is the first work of formalizing C function pointers with security guarantee.

However, accessing a checked pointer in an unchecked region violates the program partition principle of \systemname.
To overcome the discord, we develop two mechanisms in \systemname 
and maintain a stronger \textit{non-exposure} guarantee on of the non-crashing guarantee; 
i.e., no checked pointer addresses can be observed in an unchecked region.
The first mechanism is nested checked and unchecked code regions.
Users can context switch between checked and unchecked code regions 
by nested using the keywords $\echeckedtext$ $\euncheckedtext$.
The type system ensures that no checked pointers can be accessed across the context switching.
The second one is that a call to a checked pointer in unchecked code regions 
must be surrounded by a \textit{tainted shell}; 
i.e., a tainted function pointer that points to a checked regions possibly holding checked pointers.
In this case, no checked pointer address can be observed in the unchecked code regions.
In \systemname, for every checked function, 
we automatically compiler a tainted version by surrounding the function without a tainted shell.

>>>>

\section{Introduction}\label{sec:intros}

Vulnerabilities due to memory corruption, especially spatial memory corruptions, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite a large body of work that tries to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts, including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012},
have investigated ways to compile C programs 
to automatic spatial safety enforcement.
These approaches all impose performance overheads deemed too high for deployment use. 
Recently, \citet{Elliott2018} and \citet{li22checkedc} introduced and formalized \checkedc, an
open-source extension to C with new types and
annotations whose use can ensure a program’s spatial safety.
Importantly, \checkedc supports development that is 
incremental and compositional. Code regions (e.g.,
functions or whole files) designated as \emph{checked} enforce
spatial safety in a manner preserved by composition with
other checked regions. But not all regions must be checked: Checked
C's annotated \emph{checked pointers} are binary-compatible with legacy pointers, and
may coexist in the same code, which permits a deliberate (and
semi-automated) refactoring process. 
Parts of the FreeBSD kernel have
been successfully migrated to \checkedc~\cite{duanrefactoring}, and the overall performance
overhead is minimal enough for practical deployment.

@@

The guarantee \checkedc provides focuses on checked code regions,
where any stuck (i.e., ill-defined)
state reached by a well-typed program amounts to a spatial safety
violation.
Such a state can always be attributed to, i.e., 
\emph{blamed on}, the execution of code that is not in a checked core region. 
However, what are the guarantees that unchecked (unsafe) code regions can provide?
Programmers might not expect to convert every code to \checkedc,
but leave some functions in unsafe regions,
in which case the unsafe region executions should not affect other parts.
In this paper, we cover this gap by introducing \systemname,
combining \checkedc with program partitioning mechanism, 
making three main contributions.

\myparagraph{Executions Cannot Crash}
Our first contribution is a clear program partition between checked and unchecked code regions
and ensures that a \systemname program execution \textit{cannot crash} due to spatial safety violations.
We restrict the checked and unchecked pointers usage only in checked and unchecked code regions, respectively;
and develop tainted pointers for the communication between the two regions.
Heap is partitioned into two parts: a \textit{checked heap} containing all checked pointers
 and an \textit{unchecked heap} that is sandboxed.
Tainted pointers exist in the unchecked heap region and must be verified each time being used.

\myparagraph{Formalizing the Type System, Semantics and Compiler}
%
Our second contribution is a core formalism named \lang, which extends
\citet{li22checkedc} with the non-crashing guarantee and other new features below.
We prove formally the \textit{non-crashing theorem}, i.e., 
a well-typed \lang program can never crash due to spatial safety violations.
We maintain the compiler formalism with the new features extended in \systemname.
Especially, we maintain the bound-check insertions for array accesses 
without ``fat'' pointer compilation scheme, while keeping the non-crashing guarantee.
To certify the simulation relation between the \lang semantics and the compiler,
we utilize the model-based randomized testing.
This is done by a conversion tool that converts expressions from \lang into
actual \checkedc code that can be compiled by the Clang \checkedc
compiler. We create a program random generator 
based on the
typing rules of \lang and ensure that \lang and Clang Checked C are consistent after conversion, both statically and
dynamically.  
To the best of our knowledge, 
\lang is the first C(-like) language and compiler formalism with the program partitioning mechanism.

\myparagraph{Supporting Checked Pointer Callbacks and No Checked Pointer Exposure}
Our third contribution is an added-up feature
to support checked (function) pointer callbacks in unchecked code regions.
When designing a multi-threaded system, users might want 
to provide a third party interface that allows third party developers to create new program features, while keeping these programs in unchecked code regions.
Moreover, they do want to provide them a (function) pointer pointing to checked data fields.

However, accessing a checked pointer in an unchecked region violates the program partition principle of \systemname.
To resolve the conflict, we develop two mechanisms in \systemname 
and maintain a stronger \textit{non-exposure} guarantee on of the non-crashing guarantee; 
that is, no checked pointer addresses can be observed in an unchecked code region.
The first mechanism allows nested checked and unchecked code regions.
Users can context switch between checked and unchecked code regions 
by nested using the keywords $\echeckedtext$ $\euncheckedtext$.
The type system ensures that no checked pointers can be accessed across the context switching.
The second one is that a call to a checked pointer in unchecked code regions 
must be surrounded by a \textit{tainted shell}; 
i.e., a tainted function pointer that points to a checked region possibly holding checked pointers.
In this case, no checked pointer address will be observed in the unchecked code regions.
In \systemname, for every checked function, 
we automatically compile a tainted version by surrounding the function without a tainted shell.
To the best of our knowledge, \systemname is the first work of formalizing C function pointers with security guarantee.
